Constraining the Interiors of Asteroids Through Close Encounters

### 2022-03-26 paper/main.tex
Finished parameter overview

@@

>>>>

In all the above studies, we assumed a spherical planet ($J_{\ell m} = 0$ for $\ell \geq 1$). By assumption that $\mu_M \gg \mu_m$ (so that the asteroid orbit's focus is the center of mass of the central body), we have $J_{1m} = 0$. The effect of oblateness, then, is limited to the $J_{2m}$ terms, and therefore damped by a factor of $(a_M / D)^2$. We expect these parameters to have little effect on the asteroid.

Here, we define oblateness as $\epsilon = (I_z - I_x)/(\mu_M R_M^2)$, where  $I_{x,y,z}$ are the central body moments of inertia along the principal axes, and $I_x = I_y$. $R_M$ is the true radius of the body (not $a_M)$. Note that $J_{\ell m}$ is defined in equation \ref{eqn:jlm} with respect to the asteroid orbit, not the principal axes of the central body. But for an equatorial orbit, the principal axes coincide with the asteroid orbit frame and we may express $\epsilon$ simply in terms of $J_{\ell m}$ as $\epsilon = -2J_{20}$ and $J_{22} = 0$. (Some sources such as Ref.~\cite{paterLissauer2015} take this definition of $\epsilon$ as their definition of $J_{20}$). Given this conversion between $\epsilon$ and $J_{20}$, we analyze posterior uncertainty $\sigma$ of the first-order parameters as a function of $\epsilon$ across a reasonable range ofittedf oblatenesses based on those of Solar System planets \cite{paterLissauer2015}. These uncertainties are shown in figure \ref{fig:scan-oblateness}.

\begin{figure}
  \centering
  \includegraphics[width=0.88\columnwidth]{figs/oblateness.pdf}
  \vfill
  \includegraphics[width=0.88\columnwidth]{figs/oblateness-differ.pdf}
  \caption{\textit{Left}: 1- and 2-$\sigma$ confidence intervals for the first-order parameter posteriors as a function of oblateness $\epsilon$ . Linear best fit lines to $\sigma$ (black, dotted) are plotted. \textit{Right}: The deviation of posterior mean extracted from a zero-oblateness model from the true parameters given a model with oblateness $\epsilon \neq 0$. Also shown in both figures are the oblatenesses of reference Solar System bodies.}
  \label{fig:scan-oblateness}
\end{figure}

Figure \ref{fig:scan-oblateness} (top pane) demonstrates a small dependence of $\sigma$ on oblateness $\epsilon$, but the effect is far from the effect of other factors studied above. Most Solar System bodies do not reach the oblateness necessary to significantly increase precision on the first-order parameters. The linear best-fit lines match the data well, and they have slope of $(\Delta \sigma / \sigma_{\epsilon=0}) / \Delta \epsilon = -0.26$ for $\gamma_0$, $-0.94$ for $K_{22}$, and $-1.2$ for $K_{20}$. Other parameters $K_{3m}$ likely depend on oblateness similarly, but fitting these parameters is computationally more expensive.

Note that if a flyby is executed around one of the non-Earth objects noted in figure \ref{fig:scan-oblateness}, $a_M$ and $\mu_M$ will change in addition to $\epsilon$. These two parameters do affect the precision of the fit parameters, so the figure does not state that encounters with other bodies have the same precision as encounters with Earth; only that the difference in oblateness between the two bodies is of little concern.

Given the small effect of $\epsilon$ on $K_{\ell m}$, it might be tempting to neglect the effect of planetary oblateness when fitting $K_{\ell m}$ to data. However, the bottom pane of figure \ref{fig:scan-oblateness} demonstrates that this is not valid. This figure displays $K_{\ell m}$ as extracted by a fit assuming $\epsilon = 0$, but run on data generated by assuming nonzero $\epsilon$. The difference between the posterior mean $K_{\ell m}$ and true $K_{\ell m}$ are shown. Posterior uncertainties are also shown as bands. The figure shows that even for low (Moon-scale) oblateness, the fit results are inconsistent with the true $K_{\ell m}$ values, since $|\Delta K_{\ell m} = 0$ is not contained in the 2-$\sigma$ band. This effect is much worse for large oblateness, growing to a difference on the order of $10^{-2}$ between the true and fit parameters for large oblateness. Therefore, accurately modeling oblateness to high precision is essential for accurate estimation of fit parameters. For non-equatorial orbits, with $J_{22} \neq 0$, we also expect $J_{22}$ to affect the accuracy of the fit results to a similar degree, so it is also vital to use the correct asteroid orbital plane.

Note that $J_{20}$, the parameter studied in this section, has a slightly more general definition than oblateness. If the planet has a moon, the integral defining $J_{20}$ (equation \ref{eqn:jlm}) can be extended to include this extra mass, including it in $J_{20}$. Since $J_{20}$ is a second moment, this effect is magnified for large distances of the mass from the central body center of mass (though the effect is not quite quadratic because $a_m$ also increases for large distances). This process is only valid when the asteroid never passes closer to the central body than the moon.

As an order-of-magnitude estimate for this effect, two spherical masses with masses and radii of Earth and the Moon, separated by one Lunar distance, and both lying in the orbital plane has a combined $J_{20} = 0.25$. Extrapolating posterior uncertainty by the slopes of the best fit lines given earlier, this represents a decrease in $\sigma(K_{2m})$ by a factor of about one quarter.

This analysis suggests that large moons such as ours can improve fit quality, but further study of this effect is beyond the scope of this paper. Without a moon to inflate the oblateness of the central body, planetary oblateness does not significantly improve posterior uncertainty. However, correct representation of oblateness is essential to correctly estimate $K_{\ell m}$.

@@

The final encounter parameter we study is the initial rotational period of the asteroid $P_\omega$. The dynamical time scales $r_p/v_\infty$ and $\mu_m / (r_p v_\infty)$ have already been mentioned and the ratio between them and $P$ in principal may affect the posterior uncertainty $\sigma$. In figure \ref{fig:scan-period}, we show $\sigma$ as a function of $P_\omega$ for a range of periods typical of NEOs.

\begin{figure}
  \centering
  \includegraphics[height=0.89\textheight]{figs/scan-period.pdf}
  \caption{1- and 2-$\sigma$ confidence intervals for the first-order parameter posteriors (\textit{top}) and second-order parameters (\textit{bottom}) as a function of initial rotational period $P$. The reference value of 9 hr is shown as a vertical dotted line.}
  \label{fig:scan-period}
\end{figure}

Like figure \ref{fig:scan-vex}, depicting the dependence of $\sigma$ on $v_\infty$, figure \ref{fig:scan-period} shows small-scale variation in uncertainty due to the fact that varying the initial period changes the value of $\gamma$ at perigee, which affects uncertainty to a factor of about two. But a large-scale trend is also visible in many parameters. $K_{20}$ and $K_{22}$ show very large uncertainty for low $P_\omega$ because these asteroids do not tumble after perigee, and tumbling after the perigee allows better refinement of the $K_{2m}$ parameters.

We expect that quickly rotating asteroids would not tumble because, for small $P_\omega$, the dynamical variables $\bm D$, $\bm \omega$, $\alpha$, $\beta$, and which affect $\bm \tau$ vary much smaller than $\gamma$. Approximating each variable as constant over one full rotation of $\gamma$, we can integrate the first order contribution of $\bm \tau$ over $\gamma \in (0, 2\pi)$ and we see that each rotation has zero average first-order torque. Since the first-order torque over each rotational period cancels out, there is no secular torque to force the asteroid to tumble. However, this effect does not apply to the second-order parameters, since the integral over the second-order term of $\bm \tau$ does not vanish, as seen in the figure.

Another feature of figure \ref{fig:scan-period} is that $K_{\ell 0}$ is more uncertain at low $P_\omega$ than the other parameters. This is most visible in the figure for $K_{30}$. The cause of this is likely that $K_{\ell 0}$ cannot contribute to $\tau_z$ as shown in equation \ref{eqn:tidal-torque}. We already discussed that asteroids with $P_\omega$ large do not tumble, and since $\tau_x$ and $\tau_y$ are what induces tumbling, the most observable component of torque is therefore $\tau_z$, which $K_{\ell 0}$ don't affect.

The most severe effect of period on $\sigma$ is in the low-period regime ($P_\omega \lesssim 5$ hr), but in this case, the most strongly affected parameters are $K_{2m}$, which are generally known better than $K_{3m}$. The effect on the imprecise parameters $K_{3m}$ is small, except for $K_{30}$. It therefore seems as though small-period asteroids are still candidates for observation, although high-period asteroids still have better uncertainty.
